Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-06-10 document/ms.tex
comments from @benmontet

\newcommand{\project}[1]{{\sffamily #1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\terra}{\project{TERRA}}
\newcommand{\license}{MIT License}

@@

No true extra-Solar Earth analog is known.
Hundreds of planets have been found around Sun-like stars that are either
Earth-sized but on shorter periods, or else on year-long orbits but somewhat
larger.
Under strong assumptions, exoplanet catalogs have been used to make an
extrapolated estimate of the rate at which Sun-like stars host Earth analogs.
These studies are complicated by the fact that every catalog is censored by
non-trivial selection effects and detection efficiencies, and every property
(period, radius, \etc)\ is measured noisily.
Here we present a general hierarchical probabilistic framework for making
justified inferences about the population of exoplanets, taking into account
survey completeness and, for the first time, \emph{observational
uncertainties}.
We are able to make fewer assumptions about the distribution than previous
studies; we only require that the occurrence rate density be a smooth function
of period and radius (employing a Gaussian process).
By applying our method to synthetic catalogs, we demonstrate that it produces
more accurate estimates of the whole population than standard procedures based
on weighting by inverse detection efficiency.
We apply the method to an existing catalog of small planet candidates around G
dwarf stars (Petigura \etal\ 2013).
We confirm a previous result that the radius distribution changes slope near
Earth radius.
We find that the rate density of Earth analogs is about 0.02 (per star per
natural logarithmic bin in period and radius) with large uncertainty.
This number is much smaller than previous estimates made with the same data
but stronger assumptions.

@@

There are two qualitatively different methods that are commonly used to infer
the occurrence rate density from a catalog and a detection efficiency
specification.
The first is an intuitive method that we will refer to as
``inverse-detection-efficiency'' and the second is based on the likelihood
function of the catalog given a parametric rate density.
The inverse-detection-efficiency method involves making a histogram of the
objects in the catalog where each point is weighted by its inverse detection
probability.
This method is very popular in the literature (\citealt{howard, dong,
dressing, swift, petigura}) even though it is not motivated probabilistically.
The alternative likelihood method models the catalog as a Poisson realization
of the \emph{observable} rate density of exoplanets taking the survey
detection efficiencies and transit probabilities into account.
This technique has been used to constrain parametric models---a broken power
law, for example---for the occurrence rate density (\citealt{tabachnik,
youdin, dong}).
There are many reasons to prefer a result based on an optimization of a
likelihood function and as we show in \sect{model}, if we model the occurrence
rate density as a piecewise constant function, the
inverse-detection-efficiency method can be derived as a special case of the
likelihood method with a specific set of assumptions.

@@

The constraints in \eq{crazypost} can always be---and often are---propagated
as a list of $N$ samples $\{\entry_k\}^{(n)}$ from the posterior
\begin{eqnarray}\eqlabel{samples}
\{\entry_k\}^{(n)} &\sim& p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) \quad.
\end{eqnarray}
We can use these samples to approximately compute the integral in
\eq{simplemarglike} as
\begin{eqnarray}\eqlabel{importance}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &\approx&
    \frac{1}{N} \sum_{n=1}^N
    \frac{p(\{\entry_k\}^{(n)}\,|\,\ratepars)}
         {p(\{\entry_k\}^{(n)}\,|\,\interim)} \quad.
\end{eqnarray}
This is very efficient to compute as long as an evaluation of
$p(\{\entry_k\}^{(n)}\,|\,\ratepars)$ is not very expensive.
That being said, this could be a high variance estimator, depending on the
initial choice of $p(\{\entry_k\}\,|\,\interim)$.
Additionally, the support of $p(\{\entry_k\}\,|\,\ratepars)$ in $\{\entry_k\}$
space is restricted to be narrower than that of
$p(\{\entry_k\}\,|\,\interim)$.
\Eq{importance} is the \emph{importance sampling approximation} to the
integral in \eq{simplemarglike} where the trial density is the posterior
probability for the catalog measurements.

@@

Now, we can substitute \eq{poisson-like} into \eq{simplemarglike} and apply
the importance sampling approximation (\eqalt{importance}) to derive the
following expression for the marginalized likelihood
\begin{eqnarray}\eqlabel{money}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &\approx&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K
    \frac{1}{N_k} \sum_{n=1}^{N_k}
    \frac{\obs{\rate}_\ratepars (\entry_k^{(n)})}
         {p(\entry_k^{(n)}\,|\,\interim)} \quad.
\end{eqnarray}
In this equation, we're making the further assumption that the catalog treated
the objects independently.
In other words, we have per-object posterior samples
\begin{eqnarray}
\entry_k^{(n)} &\sim& p(\entry_k\,|\,\data_k,\,\interim) \quad.
\end{eqnarray}
\Eq{money} is the \emph{money equation} for our method.
It lets us efficiently compute the \emph{marginalized likelihood of the entire
set of light curves for a particular occurrence rate density}.

@@

Since \eq{ml-ide} gives the maximum likelihood solution, this result is the
minimum-variance estimator for the rate density when the completeness function
$\completeness(\entry)$ is known and the catalog uncertainties are negligible.
This is \emph{almost} the same as the inverse-detection-efficiency procedure
and if the completeness function is constant in each bin \binarea, both
methods give the same result.
If the completeness function is not piecewise-constant on the same grid as the
rate density inference, the inverse-detection-efficiency procedure will
give a higher variance estimate of the occurrence rate density than
\eq{ml-ide}.
In this \paper, we don't use either of these inverse-detection-efficiency
procedures because we are righteous.

@@

For all the results below, we run a Markov chain with $10^6$ steps for the bin
heights and update the hyperparameters every 10 steps.
We only keep the final $2 \times 10^5$ steps and discard the earlier samples
as burn-in.
The remaining samples provide an approximation to the marginalized probability
distribution for \ratepars.

@@

In \eq{gammaearth}, we define ``Earth analog'' in terms of measurable
quantities with no mention of habitability or composition.
To some readers, this might seem an unsatisfying choice.
That being said, the composition of an exoplanet is notoriously difficult to
measure even with large uncertainty and any definition of habitability is
still extremely subjective.
With this in mind, we stick to the observable definition for this \paper.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Ruth Angus (Oxford),
Jo Bovy (IAS), and
Eric Ford (PSU)
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), NASA
(grant NNX08AJ48G), and the Moore--Sloan Data Science Environment at NYU.
This research builds on ideas generated at a three-week workshop on ``Modern
Statistical and Computational Methods for Analysis of Kepler Data'' at SAMSI
in 2013 June.
This research made use of the NASA \project{Astrophysics Data System}.

@@

>>>>

\newcommand{\project}[1]{\emph{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\terra}{\project{TERRA}}
\newcommand{\license}{MIT License}

@@

No true extrasolar Earth analog is known.
Hundreds of planets have been found around Sun-like stars that are either
Earth-sized but on shorter periods, or else on year-long orbits but somewhat
larger.
Under strong assumptions, exoplanet catalogs have been used to make an
extrapolated estimate of the rate at which Sun-like stars host Earth analogs.
These studies are complicated by the fact that every catalog is censored by
non-trivial selection effects and detection efficiencies, and every property
(period, radius, \etc)\ is measured noisily.
Here we present a general hierarchical probabilistic framework for making
justified inferences about the population of exoplanets, taking into account
survey completeness and, for the first time, \emph{observational
uncertainties}.
We are able to make fewer assumptions about the distribution than previous
studies; we only require that the occurrence rate density be a smooth function
of period and radius (employing a Gaussian process).
By applying our method to synthetic catalogs, we demonstrate that it produces
more accurate estimates of the whole population than standard procedures based
on weighting by inverse detection efficiency.
We apply the method to an existing catalog of small planet candidates around G
dwarf stars (Petigura \etal\ 2013).
We confirm a previous result that the radius distribution changes slope near
Earth radius.
We find that the rate density of Earth analogs is about 0.02 (per star per
natural logarithmic bin in period and radius) with large uncertainty.
This number is much smaller than previous estimates made with the same data
but stronger assumptions.

@@

There are two qualitatively different methods that are commonly used to infer
the occurrence rate density from a catalog and a detection efficiency
specification.
The first is an intuitive method that we will refer to as
``inverse-detection-efficiency'' and the second is based on the likelihood
function of the catalog given a parametric rate density.
The inverse-detection-efficiency method involves making a histogram of the
objects in the catalog where each point is weighted by its inverse detection
probability.
This method is very popular in the literature (\citealt{howard, dong,
dressing, swift, petigura}) even though it is not motivated probabilistically.
The alternative likelihood method models the catalog as a Poisson realization
of the \emph{observable} rate density of exoplanets taking the survey
detection efficiencies and transit probabilities into account.
This technique has been used to constrain parametric models---a broken power
law, for example---for the occurrence rate density (\citealt{tabachnik,
youdin, dong}).
A result based on an optimization of a likelihood function will---under any
stated assumptions---saturate the Cram\'er--Rao bound and give an estimator of
the population with the lowest possible variance.
As we show in \sect{model}, if we model the occurrence rate density as a
piecewise constant function, the inverse-detection-efficiency method can be
derived as a special case of the likelihood method with a specific set of
assumptions.

@@

The constraints in \eq{crazypost} can always be---and often are---propagated
as a list of $N$ samples $\{\entry_k\}^{(n)}$ from the posterior
\begin{eqnarray}\eqlabel{samples}
\{\entry_k\}^{(n)} &\sim& p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) \quad.
\end{eqnarray}
We can use these samples and the Monte Carlo integral approximation to
estimate the marginalized likelihood from \eq{simplemarglike}---up to an
irrelevant constant---as
\begin{eqnarray}\eqlabel{importance}
p(\{\data_k\}\,|\,\ratepars) &\approx&
    \frac{\mathcal{Z}}{N} \sum_{n=1}^N
    \frac{p(\{\entry_k\}^{(n)}\,|\,\ratepars)}
         {p(\{\entry_k\}^{(n)}\,|\,\interim)}
\end{eqnarray}
where the constant $\mathcal{Z} = p(\{\data_k\}\,|\,\interim)$ is not a
function of the parameters.
This is very efficient to compute as long as an evaluation of
$p(\{\entry_k\}\,|\,\ratepars)$ is not very expensive.
That being said, this could be a high variance estimator, depending on the
number of independent samples $N$ and the initial choice of
$p(\{\entry_k\}\,|\,\interim)$.
Additionally, the support of $p(\{\entry_k\}\,|\,\ratepars)$ in $\{\entry_k\}$
space is restricted to be narrower than that of
$p(\{\entry_k\}\,|\,\interim)$.
Besides this caveat, in the limit of infinite samples, the approximation in
\eq{importance} becomes exact.
\Eq{importance} is the \emph{importance sampling approximation} to the
integral in \eq{simplemarglike} where the trial density is the posterior
probability for the catalog measurements.

@@

Now, we can substitute \eq{poisson-like} into \eq{simplemarglike} and apply
the importance sampling approximation (\eqalt{importance}) to derive the
following expression for the marginalized likelihood
\begin{eqnarray}\eqlabel{money}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &\approx&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K
    \frac{1}{N_k} \sum_{n=1}^{N_k}
    \frac{\obs{\rate}_\ratepars (\entry_k^{(n)})}
         {p(\entry_k^{(n)}\,|\,\interim)} \quad.
\end{eqnarray}
In this equation, we're making the further assumption that the catalog treated
the objects independently.
This is a somewhat subtle point if we were to consider targets with more than
one transiting planet---a point that we will return to below---but for the
considerations of the dataset considered in this \paper, it is a justified
simplification.
In other words, we have per-object posterior samples
\begin{eqnarray}
\entry_k^{(n)} &\sim& p(\entry_k\,|\,\data_k,\,\interim) \quad.
\end{eqnarray}
\Eq{money} is the \emph{money equation} for our method.
It lets us efficiently compute the \emph{marginalized likelihood of the entire
set of light curves for a particular occurrence rate density}.

@@

Since \eq{ml-ide} gives the maximum likelihood solution, this result is the
minimum-variance estimator for the rate density when the completeness function
$\completeness(\entry)$ is known and the catalog uncertainties are negligible.
This is \emph{almost} the same as the inverse-detection-efficiency procedure
and if the completeness function is constant in each bin \binarea, both
methods give the same result.
If the completeness function is not piecewise-constant on the same grid as the
rate density inference, the inverse-detection-efficiency procedure will
give a higher variance estimate of the occurrence rate density than
\eq{ml-ide}.
In this \paper, we don't use either of these inverse-detection-efficiency
procedures because they both fundamentally rely on the assumption of
\emph{negligible observational uncertainties}.
By applying \eq{money}, we derive the probabilistic generalization of this
model in the next section but the result is no longer analytic.

@@

For all the results below, we run a Markov chain with $10^6$ steps for the bin
heights and update the hyperparameters every 10 steps.
We only keep the final $2 \times 10^5$ steps and discard the earlier samples
as burn-in.
This results in approximately 4000 independent posterior samples estimated
using the empirical integrated autocorrelation time of the chain
(\citealt{goodman}).
These samples provide an approximation to the marginalized probability
distribution for \ratepars.

@@

In \eq{gammaearth}, we define ``Earth analog'' in terms of measurable
quantities with no mention of habitability or composition.
This might seem unsatisfying but the composition of an exoplanet is
notoriously difficult to measure even with large uncertainty and any
definition of habitability is still extremely subjective.
With this in mind, we stick to the observable definition for this \paper.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Ruth Angus (Oxford),
Jo Bovy (IAS),
Eric Ford (PSU), and
Ben Montet (Caltech/CfA)
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), NASA
(grant NNX08AJ48G), and the Moore--Sloan Data Science Environment at NYU.
This research builds on ideas generated at a three-week workshop on ``Modern
Statistical and Computational Methods for Analysis of Kepler Data'' at SAMSI
in 2013 June.
This research made use of the NASA \project{Astrophysics Data System}.

@@

\bibitem[Goodman~\&\ Weare(2010)]{goodman}
Goodman,~J. \& Weare,\ J.,
2010, Comm.\ App.\ Math.\ Comp.\ Sci., 5, 65
